Designing Qusecs Optimally: Quasi-uniform or Self-similar Layered Materials as Recursively Decomposed Solutions of Geometric Constraint Systems

### 2015-05-16 doc/GMP2015/revision_2/s02_preliminaries.tex
Revision 2. 24 hours to submission.

\section{Preliminaries}
\label{sec:prelim}

@@

Throughout this paper, a \textbf{\textsf{Theorem}}, \textbf{\textsf{Lemma}}, \textbf{\textsf{Observation}}, or \textbf{\textsf{Remark}} with an asterisk after it means that the proof appears in~\ref{sec:appendix:proofs}.

\subsection{Basic Definitions and Theory}
\label{sec:prelim:defs}
\label{sec:appendix:defs}

In this paper, a \dfn{geometric constraint system} is a multivariate polynomial (usually bilinear or quadratic) system $G(x,\delta)=0$, representing constraints with parameters $\delta$ between geometric primitives  in $\mathbb{R}^2$ represented collectively as $x\in \mathbb{R}^n$.
%
When the type of constraint (system) is fixed, the system is simply represented as $(G,\delta)$, where $G$ is the underlying constraint (hyper)graph $G = (V,E)$ with the vertices $V$ representing the geometric primitives in $\mathbb{R}^2$ and (hyper)edges $E$ representing the constraints, each with an associated parameter $\delta$.
%
For example, a \dfn{bar-joint system or linkage} $(G,\delta)$, is a graph $G=(V,E)$ with fixed length bars as edges,
% \footnote{Geometric constraint systems can also have inequalities in addition to equations, where the parameters in $\delta$ are small intervals of values rather than exact values.}
i.e. $\delta: E \rightarrow \mathbb{R}$; this represents the distance constraint system $\| x_u -x_v \|_2 = \delta_{u,v}$ for  $(u,v) \in E$, where $x_u \in \mathbb{R}^2$ represents the coordinates of $u\in V$.

In this paper, we will consider 3 types of constraint systems: bar-joint as well as \dfn{body-hyperpin} (defined formally in Section \ref{sec:bodypin}) and \dfn{pinned line-incidence} (defined formally in Section \ref{sec:pinnedline}). In all of these cases, a Cartesian \dfn{realization} or \dfn{solution} $G(p)$ of $(G,\delta)$ is an assignment of coordinates or Euclidean transformations (poses), $p: V \rightarrow \mathbb{R}^2$ or $\mathbb{R}^3$, to the vertices of $G$ satisfying the constraints with parameters $\delta$, modulo orientation preserving isometries (Euclidean rigid body motions for distance constraint).

Although the realization space itself depends on the constraint parameters $\delta$, many relevant \dfn{generic} properties of the constraint system $G(x,\delta)$ are properties of the constraint (hyper)graph $G$ and do not depend on $\delta$ (or they hold for all but a measure zero set of $\delta$ values). Many of these are properties of the Jacobian $\Delta_x G(x,\delta)$, often called the appropriate \dfn{rigidity matrix of $G$} (a matrix of indeterminates). For example, the \dfn{bar-joint rigidity matrix of the graph $G = (V,E)$} is a matrix of indeterminates representing the Jacobian of the distance map $\| x_u -x_v \|_2$  for $(u,v) \in E$. The matrix has $2$ columns per vertex in $V$ and one row per edge in $E$, where the row corresponding to edge $(u,v)$ contains the 2 coordinate indeterminates for $x_u -x_v$ (resp. $x_v-x_u$) in the 2 columns for $u$ (resp.\ $v$), that is 4 non-zero entries per row.

@@

One important property of generic constraint system or (hyper)graph\footnote{We refer to these as properties of the constraint system or as properties of the underlying (hyper)graph interchangeably.}
%that depends on the appropriate rigidity matrix of indeterminates
is \dfn{rigidity},
% (\note footnote now...),
i.e., the realizations or solutions of the corresponding constraint system being generically isolated and zero-dimensional.%(if the constraint system has equalities and exact values for the parameters $\delta$).
The result by Asimow and Roth \cite{asimow1978rigidity}  shows a constraint (hyper)graph is rigid if it is generically \dfn{infinitesimally rigid}, i.e.\ the number of independent rows of its appropriate rigidity matrix is at least the number of columns less the number of rigid body motions, which is 3 for bar-joint systems.
%This condition is called generic {\em infinitesimal rigidity}.

@@

Geometric constraint systems can also have inequalities in addition to equations, where the parameters in $\delta$ are small intervals of rather than exact values. In this case, the definition of rigidity is approximate; the solutions are isolated, small, full-dimensional connected components.

Other generic constraint system or (hyper)graph properties are the following:
% (\note we refer to these as properties of the constraint
% system or as properties of the underlying (hyper)graph
% interchangeably).
A constraint (hyper)graph $G$ is \dfn{independent} if its appropriate rigidity matrix of indeterminates has independent rows (i.e, the determinant of some square submatrix is not identically zero).
%It is {\em rigid} if the number of independent rows of the rigidity matrix is
%at least the number of columns less the number of rigid body motions,
%which is 3 for distance constraint systems.
It is \dfn{isostatic (minimally rigid, wellconstrained)} if it is both rigid and independent.
%if the number of generically independent rows or the rank of the appropriate rigidity matrix is maximal.
%For example, the maximal rank of a bar-joint rigidity matrix is $2|V| - 3$,
%where $3$ is the number of rotational and translational degrees-of-freedom of a rigid body in $\mathbb{R}^2$.
%The graph $G$ is {\em rigid} if there exists some spanning subgraph $S\subseteq G$ such that $S$ is wellconstrained.
It is \dfn{flexible} if it is not rigid, \dfn{underconstrained} if it is independent and not rigid, or \dfn{overconstrained} if it is not independent.

Defining the combinatorial independence of a subset of edges $E'\subseteq E$ to be the generic independence of corresponding rows in the appropriate rigidity matrix of indeterminates, we obtain the \dfn{rigidity matroid} of a constraint (hyper)graph $G = (V,E)$.
%The 2-dimensional {\em rigidity matroid} of a constraint (hyper)graph $G = (V,E)$ is a linear matroid  on $E$,
%where a subset of edges $E' \subseteq E$ is {\em independent} in the matroid,
%if the set of corresponding rows in the appropriate rigidity matrix of indeterminates are linearly independent.
There are various results on combinatorial characterization of independence and rigidity  and rigidity matroids for different types of graphs. For bar-joint rigidity matroid, the famous Laman's theorem \cite{laman1970graphs} states that the underlying graph is isostatic if and only if $|E| = 2|V|-3$ and $|E'| \le 2|V'|-3$ for every induced subgraph with at least 2 vertices. The result by Lovasz and Yemini \cite{lovasz1982generic}  shows that all 6-vertex-connected graphs are rigid in the plane. For bar-body rigidity matroid, Tay \cite{tay1976rigidity} proved that the underlying multigraph is isostatic if and only if it can be decomposed as $3$ edge disjoint spanning trees. White and Whiteley \cite{white1987algebraic} gave the same characterization using a different technique to study the algebraic-geometric conditions of genericity, called pure condition. Lee, Streinu and Theran \cite{lee2007graded} defined the \dfn{$(k,l)$-sparsity matroid}, where a hypergraph $G$ is called \dfn{$(k,l)$-sparse} if $|E'| \le k|V'| - l$ for any induced subgraph $(V',E')$ with at least 2 vertices, and \dfn{$(k,l)$-tight} if is $(k,l)$-sparse and $|E| = k|V| - l$. In general, given a $d$-uniform hypergraph, a $(k,l)$-sparsity condition is matroidal as long as $l \le dk-1$.

@@

%TODO
Given a bar-joint graph, a vector (of indeterminates) in the right null space of its rigidity matrix is called a \dfn{flex vector}. It has 2 entries per vertex and represents the internal motions of the system (i.e.\ modulo rigid body motions). For rigid graphs, all flex vectors are identically zero.
%
A vector (of indeterminates) in the left null space of the rigidity matrix is called a \dfn{stress vector}. It has one entry per edge and represents a \dfn{self stress} of the system. For independent graphs, all stress vectors are identically zero.
%
An \dfn{external stress} $t$ is a $1\times 2|V|$ vector  (of indeterminates) which specifies a 2D vector acting at each vertex such that the stress balance equation $sR = t$ holds, where $s$ is a vector of internal stresses, one for each bar, and $R$ is the rigidity matrix.

\subsection{Qusecs}

@@

In the remainder of this section and Sections~\ref{sec:DRP} and \ref{sec:recomb} we only consider bar-joint qusecs and graphs. Relevant formal analogies for the other 2 types of qusecs and (hyper)graphs are given in their corresponding sections on applications.

@@

\begin{figure*}\centering%
  %
  \begin{subfigure}{0.4\linewidth}\centering
    \includegraphics[height=\myMinHeight]{img/svg/triangle}
    \caption{}\label{fig:demo_graph_tri:graph}
  \end{subfigure}%
  %
  \hfill
  \begin{subfigure}{0.3\linewidth}\centering
    \includegraphics[height=\myMinHeight]{img/svg/triangle_candrp}
    \caption{}\label{fig:demo_graph_tri:candrp}
  \end{subfigure}%
  %
  \hfill
  \begin{subfigure}{0.3\linewidth}\centering
    \includegraphics[height=\myMinHeight]{img/svg/triangle_clustmindrp}
    \caption{}\label{fig:demo_graph_tri:clustmindrp}
  \end{subfigure}%
  %
  \caption{(\ref{fig:demo_graph_tri:graph}) A simple graph, $G_{tri}$, of 3 triangles intersecting trivially. (\ref{fig:demo_graph_tri:candrp}) The canonical DR-plan of $G_{tri}$. (\ref{fig:demo_graph_tri:clustmindrp}) A DR-plan of graph $G_{tri}$ which gives cluster minimality. Using the ideas from the paper~\cite{lomonosov2004graph}, the children of a triangle are the primitives of a point and an edge. Thus, the fan-in of this graph is 3, whereas the canonical/optimal DR-plan has fan-in of 2. With this counterexample, it is clear that cluster minimality is not sufficient for optimality.}
  \label{fig:demo_graph:clustmindrp}
  \label{fig:demo_graph_tri}
\end{figure*}%

\begin{definition}\label{def:drp}
    The \dfn{decomposition-recombination \mbox{(DR-)} plan} \cite{hoffman2001decompositionI} of graph $G$, $\drp{G}$, is defined as a forest that has the following properties:
    \begin{enumerate}
        \item Each node represents/contains/is a rigid subgraph of $G$.
        % \item For a node, $C$, that is  a non-trivial graph, its $N$
        % children, $C_1, \ldots, C_N$, are rigid vertex-maximal proper
        % subgraphs of $C$.
        \item Node $C$ has children $C_1,\ldots,C_N$ such that $\bigcup_{i=1}^N{C_i}=C$.
        \item A leaf node is a single vertex. A \dfn{trivial} graph is empty or a single vertex.
        \item A root node is a maximal rigid subgraph of $G$.
    \end{enumerate}
%
    % It can also be described recursively as: the root is $G$, its children are the trivial subgraphs and the DR-plans of its wellconstrained vertex-maximal proper subgraphs whose union is $G$ itself.
%
    A DR-plan is \dfn{complete} if it satisfies an additional rule: for a node $C$ that is  a non-trivial graph, its $N$ children, $C_1, \ldots, C_N$, are all of the rigid vertex-maximal proper subgraphs of $C$. This makes Rule 2 implicit, and in fact, $C$ contains all the edges in the union of its children as well. We denote a complete DR-plan of $G$ as $\comdrp{G}$.

    A DR-plan is \dfn{optimal} if it minimizes the maximum fan-in over all nodes in the tree. The maximum  fan-in is called the \dfn{size} of the DR-plan. We denote an optimal DR-plan of $G$ as $\optdrp{G}$.

    See Figures \ref{fig:demo_graph}, \ref{fig:c2c3ofk33s}, \ref{fig:demo_graph:candrpseq}, \ref{fig:demo_graph:clustmindrp}, and \ref{fig:bodypindrp} for examples of DR-plans.

%
%In general, $C$ is the graph in an arbitrary node in $CompleteDRP(G)$.
%$C_i$ is the $i^{\text{th}}$ child of $C$. It is implied that $C_i$ is
%an isostatic vertex-maximal proper subgraph.
    %Note that nodes will be referred to interchangeably as ``the node that
    %represents or contains the (sub)graph $C$'' and as simply ``$C$''.
\end{definition}
%
% \todo{This was labeled as both begin(remark) and end(observation), what was your intent?}
\begin{remark}
    For a given graph, there could be exponentially many  DR-plans and even optimal DR-plans in the size of the graph. A complete DR-plan is unique but may not be (is usually not) optimal. DR-plans of self-similar graphs are self-similar. More than one node (leaf) in a DR-plan forest may represent the same subgraph (vertex) of $G$.
\end{remark}

>>>>

\section{Preliminaries and Background}
\label{sec:prelim}

@@

We first give basic definitions and concepts in combinatorial rigidity, leading to a definition of a DR-plan, its properties, and how they relate. The section ends with a discussion of previous work on DR-plans.

\subsection{Geometric Constraint systems and Combinatorial Rigidity}
\label{sec:prelim:defs}
\label{sec:appendix:defs}

In this paper, a \dfn{geometric constraint system} is a multivariate polynomial (usually bilinear or quadratic) system $G(x,\delta)=0$, representing constraints with parameters $\delta$ between geometric primitives  in $\mathbb{R}^2$ represented collectively as $x\in \mathbb{R}^n$.
%
When the type of constraint (system) is fixed, the system is simply represented as $(G,\delta)$, where $G$ is the underlying constraint (hyper)graph $G = (V,E)$ with the vertices $V$ representing the geometric primitives in $\mathbb{R}^2$ and (hyper)edges $E$ representing the constraints, each with an associated parameter $\delta$.
%
For example, a \dfn{bar-joint system or linkage} $(G,\delta)$, is a graph $G=(V,E)$ with fixed length bars as edges,
% \footnote{Geometric constraint systems can also have inequalities in addition to equations, where the parameters in $\delta$ are small intervals of values rather than exact values.}
i.e.\ $\delta: E \rightarrow \mathbb{R}$; this represents the distance constraint system $\| x_u -x_v \|_2 = \delta_{u,v}$ for  $(u,v) \in E$, where $x_u \in \mathbb{R}^2$ represents the coordinates of $u\in V$.

In all types of geometric constraint systems we consider in this paper, a Cartesian \dfn{realization} or \dfn{solution} $G(p)$ of $(G,\delta)$ is an assignment of coordinates or Euclidean transformations (poses), $p: V \rightarrow \mathbb{R}^2$ or $\mathbb{R}^3$, to the vertices of $G$ satisfying the constraints with parameters $\delta$, modulo orientation preserving isometries (Euclidean rigid body motions).

Although the realization space itself depends on the constraint parameters $\delta$, many relevant \dfn{generic} properties of the constraint system $G(x,\delta)$ are defined to be properties of the constraint (hyper)graph $G$ and do not depend on $\delta$ (or they hold for all but a measure zero set of $\delta$ values). Many of these are properties of the Jacobian $\Delta_x G(x,\delta)$, often called the appropriate \dfn{rigidity matrix of $G$} (a matrix of indeterminates). For example, the \dfn{bar-joint rigidity matrix of the graph $G = (V,E)$} is a matrix of indeterminates representing the Jacobian of the distance map $\| x_u -x_v \|_2$  for $(u,v) \in E$. The matrix has $2$ columns per vertex in $V$ and one row per edge in $E$, where the row corresponding to edge $(u,v)$ contains the 2 coordinate indeterminates for $x_u -x_v$ (resp. $x_v-x_u$) in the 2 columns for $u$ (resp.\ $v$), i.e.\ 4 non-zero entries per row.

@@

One important property of a generic constraint system or (hyper)graph\footnote{We refer to these as properties of the constraint system or as properties of the underlying (hyper)graph interchangeably}
%that depends on the appropriate rigidity matrix of indeterminates
is \dfn{rigidity},
% (\note footnote now...),
i.e.\ the realizations or solutions of the corresponding constraint system being generically isolated and zero-dimensional.
%(if the constraint system has equalities and exact values for the parameters $\delta$).
The result by Asimow and Roth \cite{asimow1978rigidity}  shows a constraint (hyper)graph is rigid if and only if  it is generically \dfn{infinitesimally rigid}, i.e.\ the number of independent rows of its appropriate rigidity matrix is at least the number of columns less the number of rigid body motions, which is 3 for 2D bar-joint systems.
%This condition is called generic {\em infinitesimal rigidity}.

@@

Geometric constraint systems can also have inequalities in addition to equations, where the parameters in $\delta$ are small intervals rather than exact values. In this case, the definition of rigidity is approximate; the solutions are isolated, small, full-dimensional connected components.

Other generic constraint system or (hyper)graph properties are mentioned here.
% (\note we refer to these as properties of the constraint
% system or as properties of the underlying (hyper)graph
% interchangeably).
A constraint (hyper)graph $G$ is \dfn{independent} if its appropriate rigidity matrix of indeterminates has independent rows (i.e.\ the determinant of some square submatrix is not identically zero).
%It is {\em rigid} if the number of independent rows of the rigidity matrix is
%at least the number of columns less the number of rigid body motions,
%which is 3 for distance constraint systems.
It is \dfn{isostatic} (\dfn{minimally rigid}, \dfn{wellconstrained}) if it is both rigid and independent.
%if the number of generically independent rows or the rank of the appropriate rigidity matrix is maximal.
%For example, the maximal rank of a bar-joint rigidity matrix is $2|V| - 3$,
%where $3$ is the number of rotational and translational degrees-of-freedom of a rigid body in $\mathbb{R}^2$.
%The graph $G$ is {\em rigid} if there exists some spanning subgraph $S\subseteq G$ such that $S$ is wellconstrained.
It is \dfn{flexible} if it is not rigid, \dfn{underconstrained} if it is independent and not rigid, or \dfn{overconstrained} if it is not independent.

Defining the combinatorial independence of a subset of edges $E'\subseteq E$ to be the independence of corresponding rows in the rigidity matrix of indeterminates, we obtain the \dfn{rigidity matroid} of a constraint (hyper)graph $G = (V,E)$.
%The 2-dimensional {\em rigidity matroid} of a constraint (hyper)graph $G = (V,E)$ is a linear matroid  on $E$,
%where a subset of edges $E' \subseteq E$ is {\em independent} in the matroid,
%if the set of corresponding rows in the appropriate rigidity matrix of indeterminates are linearly independent.
There are various results on combinatorial characterization of independence, rigidity, and rigidity matroids for different types of (hyper)graphs. For bar-joint rigidity matroids, the famous Laman's theorem \cite{laman1970graphs} states that the underlying graph is isostatic if and only if $|E| = 2|V|-3$ and $|E'| \le 2|V'|-3$ for every induced subgraph with at least 2 vertices. The result by Lovasz and Yemini \cite{lovasz1982generic} shows that all 6-vertex-connected graphs are rigid in the plane. For bar-body rigidity matroids, Tay \cite{tay1976rigidity} proved that the underlying multigraph is isostatic if and only if it can be decomposed as $3$ edge disjoint spanning trees. White and Whiteley \cite{white1987algebraic} gave the same characterization using a different technique to study the algebraic-geometric conditions of genericity, called pure condition. Lee, Streinu and Theran \cite{lee2007graded} defined the \dfn{$(k,l)$-sparsity matroid}, where a hypergraph $G$ is called \dfn{$(k,l)$-sparse} if $|E'| \le k|V'| - l$ for any induced subgraph $(V',E')$ with at least 2 vertices, and \dfn{$(k,l)$-tight} if it is $(k,l)$-sparse and $|E| = k|V| - l$. In general, given a $d$-uniform hypergraph, a $(k,l)$-sparsity condition is matroidal as long as $l \le dk-1$.

@@

%TODO
%Given a bar-joint graph, a vector (of indeterminates) in the right null space of its rigidity matrix is called a \dfn{flex vector}. It has 2 entries per vertex and represents the internal motions of the system (i.e.\ modulo rigid body motions). For rigid graphs, all flex vectors are identically zero.
%
%A vector (of indeterminates) in the left null space of the rigidity matrix is called a \dfn{stress vector}. It has one entry per edge and represents a \dfn{self stress} of the system. For independent graphs, all stress vectors are identically zero.
%
%An \dfn{external stress} $t$ is a $1\times 2|V|$ vector  (of indeterminates) which specifies a 2D vector acting at each vertex such that the stress balance equation $sR = t$ holds, where $s$ is a vector of internal stresses, one for each bar, and $R$ is the rigidity matrix.

%\subsection{Qusecs}

@@

We note that  the remainder of this section and Sections~\ref{sec:DRP} and \ref{sec:recomb} we only consider bar-joint qusecs and graphs. Relevant formal analogies for the other 2 types of qusecs and (hyper)graphs are given along with their materials applications in Sections \ref{sec:bodypin} and \ref{sec:pinnedline}.

@@

%\begin{figure*}\centering%
%
%\begin{subfigure}{0.4\linewidth}\centering
%\includegraphics[height=\myMinHeight]{img/svg/triangle}
%    \caption{}\label{fig:demo_graph_tri:graph}
% \end{subfigure}%
%
%  \hfill
% \begin{subfigure}{0.3\linewidth}\centering
%   \includegraphics[height=\myMinHeight]{img/svg/triangle_candrp}
%    \caption{}\label{fig:demo_graph_tri:candrp}
%  \end{subfigure}%
%
%  \hfill
%  \begin{subfigure}{0.3\linewidth}\centering
%   \includegraphics[height=\myMinHeight]{img/svg/triangle_clustmindrp}
%   \caption{}\label{fig:demo_graph_tri:clustmindrp}
%  \end{subfigure}%
%
%  \caption{(\ref{fig:demo_graph_tri:graph}) A simple graph, $G_{tri}$, of 3 triangles intersecting trivially. (\ref{fig:demo_graph_tri:candrp}) The canonical DR-plan of $G_{tri}$. (\ref{fig:demo_graph_tri:clustmindrp}) A DR-plan of graph $G_{tri}$ which gives cluster minimality  (See Section \ref{sec:prev}). Using the ideas from the paper~\cite{lomonosov2004graph}, the children of a triangle are the primitives of a point and an edge. Thus, the fan-in of this graph is 3, whereas the canonical/optimal DR-plan has fan-in of 2. With this counterexample, it is clear that cluster minimality is not sufficient for optimality.}
% \label{fig:demo_graph:clustmindrp}
%  \label{fig:demo_graph_tri}
%\end{figure*}%

\begin{definition}\label{def:drp}
    The \dfn{decomposition-recombination \mbox{(DR-)} plan} \cite{hoffman2001decompositionI} of graph $G$, $\drp{G}$, is defined as a forest that has the following properties:
    \begin{enumerate}
        \item Each node represents/contains/is a rigid subgraph of $G$.
        % \item For a node, $C$, that is  a non-trivial graph, its $N$
        % children, $C_1, \ldots, C_N$, are rigid vertex-maximal proper
        % subgraphs of $C$.
        \item The children $C_1,\ldots,C_N$ of a node $C$ satisfy $\bigcup_{i=1}^N{C_i}=C$.
        \item A leaf node is a single edge. A \dfn{trivial} graph is empty or a single vertex. Note that a trivial graph is not isostatic.
        \item A root node is a vertex-maximal rigid subgraph of $G$.
    \end{enumerate}
%
    % It can also be described recursively as: the root is $G$, its children are the trivial subgraphs and the DR-plans of its wellconstrained vertex-maximal proper subgraphs whose union is $G$ itself.
%
    A DR-plan is \dfn{complete} if it satisfies an additional property: for a non-leaf node $C$, its children are all of the rigid vertex-maximal proper subgraphs of $C$. This makes Property 2 implicit.
  %and in fact, $C$ contains all the edges in the union of its children as well.
  We denote a complete DR-plan of $G$ as $\comdrp{G}$.

  A DR-plan is \dfn{optimal} if it minimizes the maximum fan-in over all nodes in the tree. The maximum  fan-in is called the \dfn{size} of the DR-plan. We denote an optimal DR-plan of $G$ as $\optdrp{G}$.
%
%In general, $C$ is the graph in an arbitrary node in $CompleteDRP(G)$.
%$C_i$ is the $i^{\text{th}}$ child of $C$. It is implied that $C_i$ is
%an isostatic vertex-maximal proper subgraph.
    %Note that nodes will be referred to interchangeably as ``the node that
    %represents or contains the (sub)graph $C$'' and as simply ``$C$''.
\end{definition}
%
% \todo{This was labeled as both begin(remark) and end(observation), what was your intent?}
\begin{remark}
More than one node (leaf) in a DR-plan forest may represent the same subgraph (vertex) of $G$.
For a given graph, there could be exponentially many DR-plans---and even optimal DR-plans---in the size of the graph. A complete DR-plan is unique but may not be (and is usually not) optimal. DR-plans of self-similar graphs are self-similar.
\end{remark}

See Figures \ref{fig:demo_graph}, \ref{fig:c2c3ofk33s}, \ref{fig:demo_graph:candrpseq}, \ref{fig:demo_graph:clustmindrp}, and \ref{fig:bodypindrp} for examples of DR-plans and how their properties relate to each other.

\subsection{Previous Work on DR-plans}
\label{sec:prev}
% \todo{We now briefly survey existing techniques for studying 2D qusecs,  many of which are \dfn{bar-joint} systems (Examples 1 and 2 above, see Sections \ref{sec:prelim}, \ref{sec:DRP}, and \ref{sec:recomb}), \dfn{body-hyperpin} systems (Example 4 and 5, see Section \ref{sec:bodypin}) or \dfn{pinned-line incidence} systems (Example 3, see Section \ref{sec:pinnedline}). The limitations of these techniques directly motivate the contributions of this paper.}
We now briefly survey existing techniques for detecting rigidity and creating DR-plans of 2D constraint systems. The limitations of these techniques directly motivate the contributions of the next  section.

\subsubsection{Finding (vertex)-maximal generically rigid subsystems}
Fast, graph-based algorithms exist (pebble-game~\cite{Jacobs:1997:PG,hoffmann1997solvablesubsets,jermann2006decomposition,Lee:2007:PGA}), for locating all maximal, \dfn{generically rigid} subsystems \seedefs. When the input itself is rigid, these algorithms do nothing, i.e.\ compute the identity function.

However, both for self-similar or just aperiodic 2D qusecs, it is imperative to recursively decompose rigid systems into their rigid subsystems, down to the level of geometric primitives, in order to understand or design properties at all scales, such as \seedefs\ \dfn{rigidity}, \dfn{flexes}, distribution of \dfn{external stresses}, boundary conditions for \dfn{isostaticity}, as well as behavior under constraint variations.

\subsubsection{Optimal Recursive Decomposition (DR-planning)}
Recursive decomposition of geometric constraint systems has been formalized \cite{hoffman2001decompositionI,hoffman2001decompositionII} and well-studied \cite{lomonosov2004graph,sitharam2005combinatorial,jermann2006decomposition} as the \dfn {Decomposition-Recombination (DR-) planning} problem \seedefsprelim. For the abovementioned classes of 2D qusecs, generic rigidity is a combinatorial property and hence each level of the decomposition should, in principle, be achievable by a graph-based algorithm \todo{rephrase? maybe not} without involving geometric information in the constraint system. Since many such decompositions can exist for a given constraint system, criteria defining desirable or optimal DR-plans and DR-planning algorithms were given in~\cite{hoffman2001decompositionI}.
%An \dfn{optimal DR-plan} is one that minimizes the \dfn{size} \seedefsprelim, i.e., the maximum number of child subsystems of any parent system. Being exponential in the size, the complexity of solving the parent constraint system is overwhelmingly dominated by the complexity of solving the child systems.

However, for overconstrained 2D qusecs, even when restricted to bar-joint systems, the optimal DR-planning problem was shown to be NP-hard \cite{lomonosov2004graph, sitharam2005combinatorial}.
The NP-hardness of the optimal DR-planning problem for 2D bar-joint graphs is partly the consequence of possibly exponential number of DR-plans. On the other hand, although the complete DR-plan is unique it could have large average fan-in and exponentially many nodes making it far from optimal.

\subsubsection{DR-plans for special classes and with other criteria}
For a special class of 2D qusecs, namely \dfn{tree-decomposable} systems~\cite{owen1991algebraic,fudos1997graph,joan-arinyo2004revisiting} common in computer aided mechanical design (which includes ruler-and-compass and Henneberg-I constructible systems), all DR-plans turn out to be optimal. This satisfies the Church-Rosser property, leading to highly efficient DR-planning algorithms. For general 2D qusecs, alternate criteria were suggested such as \dfn{cluster minimality} requiring parent systems to have a minimal set of at least 2 rigid proper subsystems as children (i.e.\ the union of no proper subset of size at least 2 child subsystems forms a rigid system); and \dfn{proper maximality}, requiring child subsystems to be maximal rigid proper subsystems of the parent system. \seedefsc.

While polynomial time algorithms were given to generate DR-plans meeting the cluster minimality criterion~\cite{lomonosov2004graph}, no such algorithm is known for the latter criterion.
%Moreover, cluster minimality does not imply optimality as shown in Figure \ref{fig:demo_graph_tri}
